---
title: "711-Daniel_D_Mondal_R_Data_Storage"
author: "Daniel D. Mondal"
date: "2025-11-17"
output: word_document
---

## 1.1 Data storage using matrix

**Question 3.3** *Create the following matrices and work out the problems that follow.*
```{r}
library(matlib)
```

1.Generate 500 observations from a standard normal distribution. Create a matrix with 50 rows and 10 columns. Name the matrix as “data”. Name the columns as x1-x10 and the rows as 1:50. Imagine that each row corresponds to a particular person and the columns are the various numerical characteristics corresponding to each person. Report the dimension of the matrix.
```{r}
x=rnorm(500)
data=matrix(x,ncol=10,byrow=F)
colnames(data)=c(paste("x",1:10,sep=""))
rownames(data)=c(1:50)
head.matrix(round(data,4))
```
2. Find the column sum, column mean and column variances.
```{r}
#column Sum:
apply(data,2,sum)
#Column Mean:
apply(data,2,mean)
#Column Variance:
apply(data,2,var)
```
3. Extract the information on x1 and x2. Compute simple correlation between x1 and x2. Produce a scatterplot with x1 and x2. Label your graph properly.
```{r}
x1_x2=data[,1:2]
#Correlation coefficient between x1 and x2
cor(x1_x2[,1],x1_x2[,2])
plot(x1_x2[,1],x1_x2[,2],xlab="X1",ylab="x2",main="Scatterplot of x1 and x2",cex=0.5)
```

4.Transform the matrix “data” to a new matrix called “data.new” whose elements are 10 + 2× elements of “data”.
```{r}
data.new=10+2*data
head.matrix(round(data.new,4))
```
5. Generate 50 observations from normal distribution with mean 10 and standard deviation 2. Store it in a vector. Append this vector with the columns of the matrix “data.new”. Call the new column as b and the new matrix as data.new1.
```{r}
x=rnorm(50,10,2)
data.new.1=cbind(data.new,"x11"=x)
head.matrix(round(data.new.1,4))
```
6. Generate 11 observations from N(10,4). These correspond to the data vector of a newly added person to the group. Append this with the rows of the matrix “data.new1”.
```{r}
s3=rnorm(11,10,2)
data.new.1=rbind(data.new.1,"51"=s3)
head.matrix(round(data.new.1,4))
```
7. Extract information on all the variables in data.new1 matrix, for any 11 individuals selected at random and without replacement. Store it in a new matrix called “A”.
```{r}
pos=sample(1:51,11,replace=F)
A=matrix(data.new.1[pos,],ncol=11,byrow=T)
head.matrix(round(A,4))
```
8. Find the dimension of A
```{r}
dim(A)
```
9.Calculate determinant of A, inverse of A and rank of the matrix A. Also find the eigen values and eigen vectors of A. From the eigen values comment on the definiteness of the matrix A.
```{r}
#Determinant of A
det(A)
#Inverse of A
head.matrix(round(solve(A),4))
#Rank of A
R(A)
eig = eigen(A)
#Eigen values
round(eig$values,4)
#Eigen Vectors
round(eig$vectors,4)
```
10. Generate 121 observations from N(10,4) and store it in a square matrix B.
```{r}
s4=rnorm(121,10,2)
B=matrix(s4,ncol=11,byrow=T)
head.matrix(round(B,4))
```
11.Perform the following matrix-algebra operations on the matrices A and B. (a) A+B (b) AT + BT (c)A × B.
```{r}
#11
# Matrix A + Matrix B
head.matrix(round(A+B,4))

# Transpose of A + Transpose of B
head.matrix(round(t(A)+t(B),4))

#Matrix Multiplication of A and B
head.matrix(round(A%*%B,4))
```
**Question 3.4.** *Suppose 12 rats were subjected to three different diets and their gains in weight (in grams) in 6 months were recorded.*

*A: 2  5  9  1  7*
*B: 3  3  1*
*C: 5  7  3  8*

*Prepare a list containing the data on A, B and C and answer the following:*
```{r}
A= c(2,5,9,1,7)
B= c(3,3,1)
C= c(5,7,3,8)

L= list(A,B,C)
L
```
(i) Display the data on B only.
```{r}
L[[2]]
```
(ii) Obtain the average gain in weight due to diet A, B and C separately.
```{r}
lapply(L,mean)
```
(iii) Obtain the standard deviation of the gain in weight due to diet A, B and C separately.
```{r}
lapply(L,sd)
```

(iv) Unlist the data and compute the grand mean.
```{r}
u = unlist(L)
mean(u)
```
```{r}
diet= rep(c("A","B","C"),times=c(5,3,4))
summary(aov(u~diet))
```
Using the Anova table:

*(v)Compute the variation in weights due to the diets.*
Sum of squares due to the diet is 20.78

*(vi)Compute within diet variation in weights.*
Sum of Squares within the diets is 62.22

*(vii) Compute the total variation in weights of rats.*
Total Sum of Squares is 83

*(viii) Is there any relationship between the answers in (v), (vi) and (vii)?*

Yes we observe that:

$${SS_{Total} = SS_{Between} + SS_{Within}}$$
and,
$$SS_{Between} + SS_{Within} = 20.78 + 62.22 = 83$$
This relationship means that the total variability observed in the rats' weight gains is decomposed into two parts:
1.Variability explained by the difference in diets ($SS_{Between}$).
2.Variability due to random error/individual differences within the same diet group ($SS_{Within}$).
