---
title: "MDTS4214_711_PS_5"
author: "Daniel D. Mondal"
date: "2026-02-25"
output: word_document
---

## Problem Set - 3 (Continued)

### 3. Problem to demonstrate the role of qualitative (ordinal) predictors in addition to quantitative predictors in multiple linear regression

Consider “diamonds” data set in R. It is in the ggplot2 package. Make a list of all the ordinal categorical variables. Identify the response.

```{r, message = F}
rm(list=ls())
library(ggplot2)
library(stargazer)
library(MASS)

attach(diamonds)
head(diamonds)
dim(diamonds)
```

(a) Run a linear regression of the response on the quality of cut. Write the fitted regression model.

```{r}
lvl=as.numeric(diamonds$cut)
code_fun=function(x){
  if(x-5==0){
    return(c(0,0,0,0))
  }else if(x-5==-1){
    return(c(1,0,0,0))
  }else if(x-5==-2){
    return(c(1,1,0,0))
  }else if(x-5==-3){
    return(c(1,1,1,0))    
  }else {
    return(c(1,1,1,1))
  }
}
pred=t(sapply(lvl,code_fun))
colnames(pred)=c("Premium","Very_Good","Good","Fair")
diamonds=cbind(diamonds,pred)
ord.model=lm(price~Premium+Very_Good+Good+Fair,data=diamonds)
summary(ord.model)
```

(b) Test whether the expected price of diamond with premium cut is significantly different from that of the ideal cut.

To Test:
$H_0: \beta_{\text{Premium}} - \beta_{\text{Ideal}} = 0$
vs
$H_1: \beta_{\text{Premium}} - \beta_{\text{Ideal}} \neq 0$

here($\beta_{\text{Ideal}}$ = $\beta_0$)

the statistic is of the form:
$t = \frac{\hat{\beta}_{\text{Premium}}}{SE(\hat{\beta}_{\text{Premium}})}$

The observed p-value for the test is $2^{-16}$, which is much smaller than our level of significance (say $\alpha = 0.05$), hence we can say that significant difference exists between the two levels.

(c) What is the expected price of a diamond of ideal cut? The Expected price of the Ideal cut is $3457.54 on an average.

(d) Modify the regression model in (a) by incorporating the predictor “table”. Write the fitted regression model.
```{r}
ord.model1=lm(price~Premium+Very_Good+Good+Fair+table,data=diamonds)
summary(ord.model1)
```

(e) Test for the significance of “table” in predicting the price of diamond.

Here we test we apply the students t-test: 

To test:$H_0: \beta_{table} = 0$ vs $H_1: \beta_{table} \neq 0$

the statistic under $H_0$: t = $\frac{\hat\beta_{table}}{SE(\beta_{table})}$

The observed p-value for the test is $2^{-16}$, which is much smaller than our level of significance (say $\alpha = 0.05$), hence we can say table is significant as a predictor.

(f) Find the average estimated price of a diamond with an average table value
and which is of fair cut.
```{r}
-6563.672+179.105*mean(table)+365.568
```

The average estimated price is $4092.765

____________________________________________________________________________________________________

## Problem Set - 5

### 1. Problem to demonstrate the utility of K nearest neighbour regression over least squares regression.

Consider a setting with n = 1000 observations. Generate:
* (i)$x_{1i}$ from $N(0, 2^2)$ and $x_{2i}$ from $Poisson(λ = 1.5)$.
* (ii)$ε_i$ from $N(0, 1)$.
* (iii)$yi = −2 + 1.4x_{1i} − 2.6x_{2i} + ε_i$

```{r}
rm(list=ls())
set.seed(123)
n=1000
x1 = rnorm(n,0,2)
x2 = rpois(n,1.5)
e  = rnorm(n)
y = -2 + 1.4*x1 - 2.6*x2 + e

data = data.frame(y,x1,x2)
head(data)
```

Split the data into train and test sets. Keep the first 800 observations as training data and the remaining as test data. Work out the following:

```{r}
train_data = data[1:800,]
test_data  = data[-(1:800),]
```

1. Fit a multiple linear regression equation of y on x1 and x2. Calculate test MSE.

```{r}
m1 = lm(y~x1+x2, data=train_data)
summary(m1)

y_test = predict(m1,newdata = test_data)

test_mse = mean((y_test-test_data[,1])^2)
test_mse
```

2. Fit a KNN model with k = 1, 2, 5, 9, 15. Calculate test MSE for each choice of k.

```{r, message = F}
library(ggplot2)
library(lattice)
library(caret)
```

```{r}
mse_knn = c()
k_choices = c(1,2,5,9,15)
for(i in 1:5)
{
  kr = knnreg(y~x1+x2, data = train_data, k= k_choices[i])
  y_knn = predict(kr,test_data)
  mse_knn[i] = mean((test_data$y - y_knn)^2)
}

output = data.frame(k_choices,mse_knn,rep(test_mse,each=5))
colnames(output) = c("k Choices","KNN Test MSE","Multiple Linear Reg Test MSE")
output
```

We consider the function:
$y_i = \frac{1}{(−2 + 1.4x_{1i} − 2.6x_{2i} + 2.9 x^2_{1i}} + 3.1sin{x_{2i}} - 1.5x_{1i}x^2_{2i} + ε_i$

Work out the problems in (1) and (2). Compare and comment on the results.

```{r}
y1 = 1/(-2 + 1.4*x1 - 2.6*x2 + 2.9*(x1^2)) + 3.1*sin(x2) - 1.5*x1*(x2^2) + e

data1 = data.frame(y1,x1,x2)
head(data1)

train_data1 = data1[1:800,]
test_data1  = data1[-(1:800),]
```

```{r}
m2 = lm(y1 ~x1+x2, data=train_data1)
summary(m2)

y1_test = predict(m2,newdata = test_data1)

test_mse1 = mean((y1_test-test_data1[,1])^2)
test_mse1
```

```{r}
mse_knn1 = c()
k_choices = c(1,2,5,9,15)
for(i in 1:5)
{
  kr1 = knnreg(y1 ~ x1+x2, data = train_data1, k= k_choices[i])
  y1_knn = predict(kr1,test_data1)
  mse_knn1[i] = mean((test_data1[,1] - y1_knn)^2)
}

output = data.frame(k_choices,mse_knn1,rep(test_mse1,each=5))
colnames(output) = c("k Choices","KNN Test MSE","Multiple Linear Reg Test MSE")
output
```

* From the above comparsion we see that, KNN method performs much better than multiple linear regression. KNN produces a much lower test mean square error as the model is non-linear, hence multiple linear regression performs poorly. 
* We observe that KNN model for k = 1 is minimum, but we see that the increment in MSE for different values of k is substantially small hence we can opt for a moderately high to high value of k.