---
title: "MDTS4214_711_PS_8"
author: "Daniel D. Mondal"
date: "2026-03-30"
output: word_document
---

## 1. Problem to demonstrate the fitting of a binary response variable using logistic regression

### a. Consider the Weekly data available in the ISLR package of R. Use the data from 1990 to 2008 as train data and the data on 2009-2010 as test data.

```{r}
rm(list=ls())
library(ISLR)
library(caret)

Direction = ifelse(Weekly$Direction=="Up",1,0)
data = cbind(Weekly[,-9], Direction)
attach(data)
head(data)
str(data)

train_data = data[Year<2009,]
test_data = data[Year>=2009,]
```

### b. Using the training data, fit a logistic regression with Direction as the response and the five lag variables plus Volume as predictors. Are there any significant predictors? If not, which of the tests is reporting the least p value. Interpret the coefficient corresponding to that test.
```{r}
model1 =  glm(Direction ~Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = train_data, 
              family = binomial(link = "logit"))
summary(model1)

exp(-0.06231)
```

**Interpretation:**

- Significant Predictors: Based on the standard $\alpha = 0.05$ level, none of the predictors are statistically significant in the training set.
- Least p-value: Typically, Lag2 shows the lowest p-value (often around 0.03 in the full dataset.
- Coefficient Interpretation: If we look at the coefficient for Lag2 ($\hat{\beta}_{Lag1}$), a negative value indicates that for every one-unit increase in the return ome days ago, the log-odds of the market going "Up" today decreases by that 0.0628 units, holding other variables constant.

### c. Estimate the odds of market going “up” at median values of the predictors.
```{r}
median_pred =apply(train_data[,2:7], 2, median)
median_pred

#required in data frame format
median_pred = data.frame(t(apply(train_data[,2:7], 2, median)))
median_pred
log_odds = predict(model1, newdata = median_pred)
odds = exp(log_odds)
odds
```

To understand the actual count of the directions
```{r}
table(test_data$Direction)
```

### e. Using the fitted model and the predictor values from the test data, find an optimal cut point that minimizes TPR(1 − FPR). Report the confusion matrix corresponding to the optimal cut-point value. From the confusion matrix compute the test error rate.

```{r, warning=F}
pred_test = predict(model1, newdata = test_data[,2:7], type = "response")

tpr = c()
fpr = c()
measure = c()
for(i in 1:length(pred_test))
{
  pred_Direction = ifelse(pred_test < pred_test[i],0,1)
  
  conf_matrix = confusionMatrix(as.factor(pred_Direction),as.factor(test_data$Direction))
  tpr[i] = conf_matrix$byClass["Sensitivity"]
  fpr[i] = 1 - conf_matrix$byClass["Specificity"] 
  measure[i] = tpr[i]*(1-fpr[i])
}

output = data.frame("Cut_Point" = pred_test, "TPR" = tpr,"FPR" = fpr,"Compare" = measure)
head(output)

output[which.max(output$Compare),]
opt_cutpoint = output[which.max(output$Compare),1]
cat("The cutpoint for which TPR(1-FPR) is minimum:", opt_cutpoint)



opt_pred = ifelse(pred_test < opt_cutpoint, 0,1)
opt_conf_matrix = confusionMatrix(as.factor(opt_pred),as.factor(test_data$Direction))
opt_conf_matrix
```

We obtain the error measures from the confusion matrix.

### f. Is there any change in the test error rate if we have worked with only the two predictors lag 1 and lag 2?
```{r,warning=F}
model2 = glm(Direction ~ Lag1 + Lag2, data = train_data, family = binomial)
prob = predict(model2, newdata = test_data, type = "response")

tpr1 = c()
fpr1 = c()
measure1 = c()
for(i in 1:length(prob))
{
  predict_reduced = ifelse(prob < prob[i],0,1)
  
  conf_matrix1 = confusionMatrix(as.factor(predict_reduced),as.factor(test_data$Direction))
  tpr1[i] = conf_matrix1$byClass["Sensitivity"]
  fpr1[i] = 1 - conf_matrix1$byClass["Specificity"] 
  measure1[i] = tpr1[i]*(1-fpr1[i])
}

output1 = data.frame("Cut_Point" = prob, "TPR" = tpr1,"FPR" = fpr1,"Compare" = measure1)
head(output1)

output1[which.max(output$Compare),]
opt_cutpoint1 = output1[which.max(output$Compare),1]
cat("The cutpoint for which TPR(1-FPR) is minimum:", opt_cutpoint1)



opt_pred = ifelse(prob < opt_cutpoint1, 0,1)
opt_conf_matrix = confusionMatrix(as.factor(opt_pred),as.factor(test_data$Direction))
opt_conf_matrix
```

### Evaluation of Model Performance

To assess the efficacy of our classification models:

- Sensitivity (Recall): Measures the proportion of actual "Up" days that were correctly identified by the model.
$$\text{Sensitivity} = \frac{TP}{TP + FN}$$

- Specificity: Measures the proportion of actual "Down" days that were correctly identified.
$$\text{Specificity} = \frac{TN}{TN + FP}$$

Comparative Analysis:

- Model 1: Full Feature SetThe full model provides a balanced predictive profile. With a sensitivity of $0.5116$ and a specificity of $0.5574$, the model demonstrates a relatively even capability in detecting both market advances and declines.

- Model 2: Reduced Feature Set ($Lag1$ & $Lag2$)The streamlined model shows a shift in classification priority. While it achieves a notably higher specificity of $0.7869$, making it quite effective at identifying "Down" days, this gain comes at the expense of sensitivity, which drops to $0.4419$.
