---
title: "MDTS4214_711_PS_8"
author: "Daniel D. Mondal"
date: "2026-03-30"
output: word_document
---

## 1. Problem to demonstrate the fitting of a binary response variable using logistic regression

### a. Consider the Weekly data available in the ISLR package of R. Use the data from 1990 to 2008 as train data and the data on 2009-2010 as test data.

```{r}
rm(list=ls())
library(ISLR)
library(caret)

Direction = ifelse(Weekly$Direction=="Up",1,0)
data = cbind(Weekly[,-9], Direction)
attach(data)
head(data)
str(data)

train_data = data[Year<2009,]
test_data = data[Year>=2009,]
```

### b. Using the training data, fit a logistic regression with Direction as the response and the five lag variables plus Volume as predictors. Are there any significant predictors? If not, which of the tests is reporting the least p value. Interpret the coefficient corresponding to that test.
```{r}
model1 =  glm(Direction ~Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = train_data, 
              family = binomial(link = "logit"))
summary(model1)

exp(-0.06231)
```

**Interpretation:**

- Significant Predictors: Based on the standard $\alpha = 0.05$ level, none of the predictors are statistically significant in the training set.
- Least p-value: Typically, Lag2 shows the lowest p-value (often around 0.03 in the full dataset.
- Coefficient Interpretation: If we look at the coefficient for Lag2 ($\hat{\beta}_{Lag1}$), a negative value indicates that for every one-unit increase in the return ome days ago, the log-odds of the market going "Up" today decreases by that 0.0628 units, holding other variables constant.

### c. Estimate the odds of market going “up” at median values of the predictors.
```{r}
median_pred =apply(train_data[,2:7], 2, median)
median_pred

#required in data frame format
median_pred = data.frame(t(apply(train_data[,2:7], 2, median)))
median_pred
log_odds = predict(model1, newdata = median_pred)
odds = exp(log_odds)
odds
```

To understand the actual count of the directions
```{r}
table(test_data$Direction)
```

### e. Using the fitted model and the predictor values from the test data, find an optimal cut point that minimizes TPR(1 − FPR). Report the confusion matrix corresponding to the optimal cut-point value. From the confusion matrix compute the test error rate.

```{r, warning=F}
pred_test = predict(model1, newdata = test_data[,2:7], type = "response")

tpr = c()
fpr = c()
measure = c()
for(i in 1:length(pred_test))
{
  pred_Direction = ifelse(pred_test < pred_test[i],0,1)
  
  conf_matrix = confusionMatrix(as.factor(pred_Direction),as.factor(test_data$Direction))
  tpr[i] = conf_matrix$byClass["Sensitivity"]
  fpr[i] = 1 - conf_matrix$byClass["Specificity"] 
  measure[i] = tpr[i]*(1-fpr[i])
}

output = data.frame("Cut_Point" = pred_test, "TPR" = tpr,"FPR" = fpr,"Compare" = measure)
head(output)

output[which.max(output$Compare),]
opt_cutpoint = output[which.max(output$Compare),1]
cat("The cutpoint for which TPR(1-FPR) is minimum:", opt_cutpoint)



opt_pred = ifelse(pred_test < opt_cutpoint, 0,1)
opt_conf_matrix = confusionMatrix(as.factor(opt_pred),as.factor(test_data$Direction))
opt_conf_matrix
```

We obtain the error measures from the confusion matrix.

### f. Is there any change in the test error rate if we have worked with only the two predictors lag 1 and lag 2?
```{r,warning=F}
model2 = glm(Direction ~ Lag1 + Lag2, data = train_data, family = binomial)
prob_reduced = predict(model2, newdata = test_data, type = "response")

tpr1 = c()
fpr1 = c()
measure1 = c()
for(i in 1:length(prob_reduced))
{
  pred_reduced = ifelse(prob_reduced < prob_reduced[i],0,1)
  
  conf_matrix = confusionMatrix(as.factor(pred_reduced),as.factor(test_data$Direction))
  tpr1[i] = conf_matrix$byClass["Sensitivity"]
  fpr1[i] = 1 - conf_matrix$byClass["Specificity"] 
  measure1[i] = tpr1[i]*(1-fpr1[i])
}

output1 = data.frame("Cut_Point" = pred_test, "TPR" = tpr1,"FPR" = fpr1,"Compare" = measure1)
head(output1)

output1[which.max(output$Compare),]
opt_cutpoint1 = output1[which.max(output$Compare),1]
cat("The cutpoint for which TPR(1-FPR) is minimum:", opt_cutpoint1)



opt_pred = ifelse(prob_reduced < opt_cutpoint1, 0,1)
opt_conf_matrix = confusionMatrix(as.factor(opt_pred),as.factor(test_data$Direction))
opt_conf_matrix
```

*Interpretation:* 

In classification, we define:
- Sensitivity (True Positive Rate, TPR): The ability of the model to correctly identify "Up" days.

$$\text{Sensitivity} = \frac{TP}{TP + FN}$$

- Specificity (True Negative Rate, TNR): The ability of the model to correctly identify "Down" days.

$$\text{Specificity} = \frac{TN}{TN + FP}$$

- Model 1 (Full Model):This model demonstrates a more balanced predictive performance. With moderate values for both Sensitivity and Specificity, it shows a consistent ability to identify both "Up" and "Down" market movements, suggesting it captures a broader range of the underlying data variance, even if the individual predictors are not highly significant.

- Model 2 (Reduced Model - Lag1 & Lag2):The reduced model exhibits a significant bias toward one class. While it achieves a high Specificity, correctly identifying the "Down" days, it suffers from a negligible Sensitivity. 

$$\text{Sensitivity} \approx 0 \implies TP \approx 0$$

This indicates that the model is effectively predicting the majority class or failing to find a boundary that captures "Up" movements. While the test error rate might be lower due to the high Specificity, the model lacks practical utility for predicting positive market shifts.